### 2023-11-12 sections/2-methods.tex
fixing error in repeated rnn explanation
removed scaling behavior of radii section

\subsubsection {Scaling Behavior of Cluster Radii}
\label{subsubsec:methods:clustering:guaranteed-decrease-in-cluster-radii}

While it may be tempting to assume that cluster radii decrease with each application of Partition (refer to Algorithm \ref{alg:methods:clustering:partition}), unfortunately, this assumption is incorrect. 
Fortunately, we \emph{can} make some guarantees about the scaling behavior of cluster radii;
in particular, we prove in this section that cluster radii will stop increasing after at most $d$ partitions, where $d$ is the fractal dimensionality of the dataset.

We can describe a $d$-dimensional distribution of data by choosing some set of $d$ mutually orthogonal axes.
Let $2R$ be the maximum pairwise distance among the points in the dataset. 
We choose the axes such that the two points that are $2R$ apart lie \emph{along one of the axes}. 
Thus, a $d$-dimensional hyper-sphere of radius $R$ would bound the dataset. 
In the worst case, (i.e., with a uniform-density distribution that fills the $d$-sphere), our axes will be such that $2R$ is the maximum pairwise distance \emph{along every axis}. 
Such a distribution would also produce a balanced cluster tree.

In this case, Partition will select a maximally distant pair of points to use as poles;
without loss of generality, say it chooses two points at the extrema one of the $d$ axes. 
After one partition, the maximum pairwise distance along that axis will be halved, i.e. bounded above by $R$.
The next recursive Partition will select another of the $d$ axes, and once again, the distance along that axis will be bounded above by $R$
Thus, after at most $d$ partitions, the maximum pairwise distance along each axis will be bounded above by $R$. 
The overall (i.e., not restricted to one axis) maximum pairwise distance 
will be bounded above by $R\sqrt{2}$ by, for example, two instances that lie at the extrema of different axes. 
See Figure~\ref{fig:methods:scaling-behavior} for an example.

Thus, starting with a cluster $C$ of radius $R$, after at most $d$ Partitions, the descendants of $C$ will each have radius
bounded above by $\frac{R}{\sqrt{2}}$. In other words, cluster radii are guaranteed to decrease by a multiplicative factor of $\frac{\sqrt{2}}{2}$ after at 
most $d$ partitions.

Note that, in practice, we never see a balanced clustering.
Partition produces unbalanced trees due to the varying density of the sampling in different regions of the manifold and the low dimensional ``shape'' of the manifold.
Further, in practice, the cluster radii decrease by a factor much larger than $\frac{\sqrt{2}}{2}$ with almost every partition;
the upper bound of $d$ partitions is seldom realized.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=3.4in]{images/geometry/geometry.pdf}
    \caption{
        Scaling behavior of radii on a uniform-density disk, which represents the worst-case scenario for a two-dimensional distribution. 
        $R_0$ denotes the radius of the root cluster $C_0$, and $o_0$ denotes its center. 
        After one application of Algorithm~\ref{alg:methods:clustering:partition}, we have $C_1$, with radius $R_1$ and center $o_1$.
        The right triangle formed by $o_0$, $o_1$, and $y_+$ in $C_1$ shows that $R_0 < R_1$.
        Hence, the radius of a child cluster can be larger than its parent.
        However, after another application of Algorithm~\ref{alg:methods:clustering:partition}, we have consumed all $d$ orthogonal axes, 
        as shown in $C_2$.
        Now, clearly $R_2 < R_0$.
        In fact, for each subsequent application of Algorithm~\ref{alg:methods:clustering:partition}, the radius of the resulting cluster is bounded above by $\frac{\sqrt{2}R}{2}$.
    }
    \label{fig:methods:scaling-behavior}
\end{figure}

@@

where $\mu$ is the harmonic mean of the local fractal dimensions (abbreviated lfd in algorithm below) of each cluster which contains one of the nearest neighbors found thus far.
We use the harmonic mean to ensure that $\mu$ is not dominated by outlier clusters with very high fractal dimension. 
Intuitively, the factor by which we increase the radius should be \emph{inversely} related to the number of points found so far. 
Additionally, when the local fractal dimension at the radius scale from the previous iteration is low, this suggests that the data are relatively concentrated in that region.
In other words, a small increase in the radius would likely encounter unoccupied space around the manifold, so a larger radial increase is needed.
Thus, the factor of radius increase should be \emph{directly} related to the local fractal dimension.
Once $|H| >= k$, we sort the points by their distance to the query and return the $k$ closest points.

>>>>

% % WE WILL USE THIS FOR COMPRESSION, so removing it for now

% \subsubsection {Scaling Behavior of Cluster Radii}
% \label{subsubsec:methods:clustering:guaranteed-decrease-in-cluster-radii}

% While it may be tempting to assume that cluster radii decrease with each application of Partition (refer to Algorithm \ref{alg:methods:clustering:partition}), unfortunately, this assumption is incorrect. 
% Fortunately, we \emph{can} make some guarantees about the scaling behavior of cluster radii;
% in particular, we prove in this section that cluster radii will stop increasing after at most $d$ partitions, where $d$ is the fractal dimensionality of the dataset.

% We can describe a $d$-dimensional distribution of data by choosing some set of $d$ mutually orthogonal axes.
% Let $2R$ be the maximum pairwise distance among the points in the dataset. 
% We choose the axes such that the two points that are $2R$ apart lie \emph{along one of the axes}. 
% Thus, a $d$-dimensional hyper-sphere of radius $R$ would bound the dataset. 
% In the worst case, (i.e., with a uniform-density distribution that fills the $d$-sphere), our axes will be such that $2R$ is the maximum pairwise distance \emph{along every axis}. 
% Such a distribution would also produce a balanced cluster tree.

% In this case, Partition will select a maximally distant pair of points to use as poles;
% without loss of generality, say it chooses two points at the extrema one of the $d$ axes. 
% After one partition, the maximum pairwise distance along that axis will be halved, i.e. bounded above by $R$.
% The next recursive Partition will select another of the $d$ axes, and once again, the distance along that axis will be bounded above by $R$
% Thus, after at most $d$ partitions, the maximum pairwise distance along each axis will be bounded above by $R$. 
% The overall (i.e., not restricted to one axis) maximum pairwise distance 
% will be bounded above by $R\sqrt{2}$ by, for example, two instances that lie at the extrema of different axes. 
% See Figure~\ref{fig:methods:scaling-behavior} for an example.

% Thus, starting with a cluster $C$ of radius $R$, after at most $d$ Partitions, the descendants of $C$ will each have radius
% bounded above by $\frac{R}{\sqrt{2}}$. In other words, cluster radii are guaranteed to decrease by a multiplicative factor of $\frac{\sqrt{2}}{2}$ after at 
% most $d$ partitions.

% Note that, in practice, we never see a balanced clustering.
% Partition produces unbalanced trees due to the varying density of the sampling in different regions of the manifold and the low dimensional ``shape'' of the manifold.
% Further, in practice, the cluster radii decrease by a factor much larger than $\frac{\sqrt{2}}{2}$ with almost every partition;
% the upper bound of $d$ partitions is seldom realized.

% \begin{figure}[ht!]
%     \centering
%     \includegraphics[width=3.4in]{images/geometry/geometry.pdf}
%     \caption{
%         Scaling behavior of radii on a uniform-density disk, which represents the worst-case scenario for a two-dimensional distribution. 
%         $R_0$ denotes the radius of the root cluster $C_0$, and $o_0$ denotes its center. 
%         After one application of Algorithm~\ref{alg:methods:clustering:partition}, we have $C_1$, with radius $R_1$ and center $o_1$.
%         The right triangle formed by $o_0$, $o_1$, and $y_+$ in $C_1$ shows that $R_0 < R_1$.
%         Hence, the radius of a child cluster can be larger than its parent.
%         However, after another application of Algorithm~\ref{alg:methods:clustering:partition}, we have consumed all $d$ orthogonal axes, 
%         as shown in $C_2$.
%         Now, clearly $R_2 < R_0$.
%         In fact, for each subsequent application of Algorithm~\ref{alg:methods:clustering:partition}, the radius of the resulting cluster is bounded above by $\frac{\sqrt{2}R}{2}$.
%     }
%     \label{fig:methods:scaling-behavior}
% \end{figure}

@@

where $\mu$ is the harmonic mean of the local fractal dimensions (abbreviated lfd in algorithm below) of each cluster which contains one of the nearest neighbors found thus far.
We use the harmonic mean to ensure that $\mu$ is not dominated by outlier clusters with very high fractal dimension. 
Intuitively, the factor by which we increase the radius should be \emph{inversely} related to the number of points found so far. 
Additionally, when the local fractal dimension at the radius scale from the previous iteration is low, this suggests that the data are relatively concentrated in that region.
In other words, a small increase in the radius would likely encounter unoccupied space around the manifold, so a larger radial increase is needed.
Thus, the factor of radius increase should be \emph{inversely} related to the local fractal dimension.
Once $|H| >= k$, we sort the points by their distance to the query and return the $k$ closest points.
